A Responsible Machine Learning Workflow\\{with Focus on Interpretable Models, Post-hoc Explanation, and Discrimination Testing

### 2020-02-26 article-information-2019.tex
edits after mdpi edits ...

@@

%=================================================================
\documentclass[information,article,submit,moreauthors,pdftex]{definitions/mdpi}

%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{xx}
\issuenum{1}
\articlenumber{5}
\pubyear{2019}
\copyrightyear{2019}
%\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}
%\updates{yes} % If there is an update available, un-comment this line

%=================================================================
% Full title of the paper (Capitalized)
\Title{A Responsible Machine Learning Workflow\\\normalsize{With Focus on Interpretable Models, Post-hoc Explanation, and Discrimination Testing}}

% Authors, for the paper (add full first names)
\Author{Navdeep Gill $^{1, \dagger}$, Patrick Hall $^{1,3,\dagger,*}$, Kim Montgomery $^{1,\dagger}$, and Nicholas Schmidt $^{2, \dagger, *}$}

% Authors, for metadata in PDF
\AuthorNames{Navdeep Gill, Patrick Hall, Kim Montgomery, and Nicholas Schmidt}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad H2O.ai\\
$^{2}$ \quad BLDS, LLC\\
$^{3}$ \quad The George Washington University}

% Contact information of the corresponding author
\corres{Correspondence: phall@h2o.ai; nschmidt@bldsllc.com}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBMs) and explainable neural networks (XNNs), a deep learning architecture well-suited for structured data, are assessed on simulated data and publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and with global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using measures with long-standing legal precedents, adverse impact ratio (AIR), marginal effect (ME), and standardized mean difference (SMD), along with straightforward group fairness measures. By combining interpretable models, post-hoc explanations, and discrimination testing with accessible software tools, this text aims to provide a template workflow for important ML applications that require high accuracy and interpretability and that mitigate risks of discrimination.}

% Keywords
\keyword{Deep Learning; Disparate Impact; Explanation; Fairness; Gradient Boosting Machine; Interpretable; Machine Learning; Neural Network; Python}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{section}{-1}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/gx/en/issues/data-and-analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}.} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is becoming clear. ML models, a common application of AI, can present serious risks. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{\textit{When a Computer Program Keeps You in Jail}}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}, and ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. This manuscript makes no claim that these interdependent issues of ML opaqueness, discrimination, privacy harms, and security vulnerabilities have been resolved, even as singular entities, and much less as complex intersectional phenomena. However, Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, mostly in the form of interpretable models, post-hoc explanation, and DI and discrimination testing, that responsible practitioners can use to address a subset of these vexing problems.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve Supervisory and Regulatory (SR) Letter 11-7, and the European Union (EU) General Data Protection Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

Section \ref{sec:m_and_m} describes methods and materials, including training datasets, interpretable and constrained models, post-hoc explanations, tests for DI and other social discrimination, and public and open source software resources associated with this text. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models, and post-hoc explanation and discrimination testing results are also presented for interpretable models. Of course, an even wider array of tools and techniques are likely helpful to fully minimize discrimination, inaccuracy, privacy, and security risks associated with ML models. Section \ref{sec:disc} puts forward a more holistic responsible ML modeling workflow, and addresses the burgeoning Python ecosystem for responsible AI, along with appeal and override of automated decisions, and discrimination testing and remediation in practice. Section \ref{sec:con} closes this manuscript with a brief summary of the outlined methods, materials, results, and discussion.

@@

The simulated data (see Subsection \ref{ssec:sim_data}) is based on the well-known Friedman datasets. Its known feature importance and augmented discrimination characteristics are used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1979tree}, \cite{friedman1991multivariate}. The mortgage data (see Subsection \ref{ssec:mort_data}) is sourced from the Home Mortgage Disclosure Act (HMDA) database and is a fairly realistic data source for demonstrating the template workflow.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} To provide a sense of fit differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. Because the unconstrained ML models, gradient boosting machines (GBMs, e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs, e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), do not exhibit convincing accuracy benefits on the simulated or mortgage data and can also present the unmitigated risks discussed above, further explanation and discrimination analyses are applied only to the constrained, interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} (See Subsection \ref{ssec:mgbm}) and XNNs (see Subsection \ref{ssec:xnn}, \cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analyses. MGBM and XNN interpretable model architectures are selected for the example workflow because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated to also evaluate the constrained versions of these models.

The same can be said of the selected explanation methods and discrimination tests. Due to their post-hoc nature, they can often be shoe-horned into existing ML workflows and pipelines. Presented explanation techniques include PD and ICE (see Subsection \ref{ssec:pd_ice}) and Shapley values (see Subsection \ref{ssec:shap}) \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discrimination testing methods discussed (see Subsection \ref{ssec:di}) include AIR, ME, and SMD \cite{feldman2015certifying}, \cite{cohen1988statistical}, \cite{cohen1992power}.\footnote{See: Part 1607 - Uniform Guidelines on Employee Selection Procedures (1978) \textsection 1607.4.} Accuracy and other confusion matrix measures are also reported by demographic segment \cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available on GitHub (see Subsection \ref{ssec:soft}).

@@

Simulated data is created based on a signal-generating function, $f$, applied to input data, $\mathbf{X}$, first proposed in \citet{friedman1979tree} and extended in \citet{friedman1991multivariate}:

\begin{equation}
\begin{aligned}\label{eq:sim}\small
f(\mathbf{X}) = 10\sin(\pi~X_{\text{Friedman},1}~X_{\text{Friedman},2}) + 20(X_{\text{Friedman},3} - 0.5)^2 + 10~X_{\text{Friedman},4} + 5~X_{\text{Friedman},5}
\end{aligned}
\end{equation}

\noindent where each $X_{\text{Friedman},j}$ is a random uniform feature in $[0,1]$. In Friedman's texts, a Gaussian noise term was added to create a continuous output feature for testing spline regression methodologies. In this manuscript, the signal-generating function and input features are modified in several ways.  Two binary features, a categorical feature with five discrete levels, and a bias term are introduced into $f$ to add a degree of complexity that may more closely mimic real-world settings. 
%When generating the data, random draws for each observation are constructed such that the features can be correlated with each other, some highly so. For example, X1 and X3 have a Pearson correlation coefficient equal to 0.58.
For binary classification analysis, the Gaussian noise term is replaced with noise drawn from a logistic distribution and coefficients are re-scaled to be $\frac{1}{5}$ of the size of those used by Friedman, and any $f(\mathbf{X})$ value above 0 is classified as a positive outcome, while $f(\mathbf{X})$ values less than or equal to zero are designated as negative outcomes.  Finally, $f$ is augmented with two hypothetical protected class-control features with known dependencies on the binary outcome to allow for discrimination testing. The simulated data is generated to have eight input features, twelve after numeric encoding of categorical features, and a binary outcome, two class-control features, and 100,000 instances. The simulated data is then split into a training and test set, with 80,000 and 20,000 instances, respectively.  Within the training set, a 5-fold cross-validation indicator is used for training all models. For an exact specification of the simulated data, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

The mortgage dataset analyzed here is a random sample of consumer-anonymized loans from the HDMA database. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e., the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  After data cleaning and preprocessing to encode categorical features and create missing markers, the mortgage data contains ten input features and the binary outcome, \textit{high-priced}. The data is split into a training set with 160,338 loans and a marker for 5-fold cross-validation and a test set containing 39,662 loans. While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (loan to value (LTV) ratio, debt to income (DTI) ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider. See the resources put forward in Section \ref{ssec:soft} and Appendix \ref{a_sec:mort} for more information regarding the HMDA mortgage data.

@@

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints, with respect to each input feature, $X_j$, and a target feature, $\mathbf{y}$, independently. An MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, and each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$. For some instance, $\mathbf{x}$, a trained MGBM model, $g^{\text{MGBM}}$, takes the form:

@@

\noindent As in unconstrained GBM, $\Theta^{\text{mono}}_b$ is selected in a greedy, additive fashion by minimizing a regularized loss function that considers known target labels, the predictions of all subsequently trained trees in $g^{\text{MGBM}}$, and the $b$-th tree splits applied to $\mathbf{x}$, $T_b(\mathbf{x};\Theta^\text{mono}_b)$, in a numeric loss function (e.g., squared loss, Huber loss), and a regularization term that penalizes complexity in the current tree. See Appendices \ref{a_ssec:not} and \ref{a_ssec:mgbm} for details pertaining to MGBM training.

Herein, two $g^\text{MGBM}$ models are trained. One on the simulated data and one on the mortgage data. In both cases, positive and negative monotonic constraints for each $X_j$ are selected using domain knowledge, random grid search is used to determine other hyperparameters, and 5-fold cross-validation and test partitions are used for model assessment. For exact parameterization of the two $g^\text{MGBM}$ models, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bear a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}Ms), which consider main effects and a small number of 2-way interactions and may also incorporate boosting into their training \cite{esl}, \cite{ga2m}. XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. A trained XNN function, $g^{\text{XNN}}$, applied to some instance , $\mathbf{x}$, is defined as:

\begin{equation}
\begin{aligned}
\label{eq:xnn}
g^{\text{XNN}}(\mathbf{x}) = \mu_0 + \sum_{k=0}^{K-1}\gamma_k n_k(\sum^{J-1}_{j=0}\beta_{k, j}x_j)
\end{aligned}
\end{equation}

\noindent where $\mu_0$ is a global bias for $K$ individually specified ANN subnetworks, $n_k$, with weights $\gamma_k$. The inputs to each $n_k$ are themselves a linear combination of the $J$ modeling inputs and their associated $\beta_{k,j}$ coefficients in the deepest, i.e., \textit{projection}, layer of $g^\text{XNN}$.

Two $g^\text{XNN}$ models are trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in 5 training folds and in a test data partition. $L_1$ regularization is applied to network weights to induce a sparse and interpretable model, where each $n_k$ and corresponding $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft} and Appendices  \ref{a_ssec:not} and \ref{a_ssec:xnn}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One-dimensional Partial Dependence and Individual Conditional Expectation}\label{ssec:pd_ice}

>>>>

% CC by 4.0

@@

%=================================================================
%%%\documentclass[information,article,submit,moreauthors,pdftex]{definitions/mdpi} 
\documentclass[information,article,accept,moreauthors,pdftex]{Definitions/mdpi}
\usepackage{soul}
%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{xx}
\issuenum{1}
\articlenumber{5}
\pubyear{2020}
\copyrightyear{2020}
%\externaleditor{Academic Editor: name}
\history{Received: 21 December 2019; Accepted: 25 February 2020; Published: date}
%\updates{yes} % If there is an update available, un-comment this line

%=================================================================
% Full title of the paper (Capitalized)
\Title{A Responsible Machine Learning Workflow\\{with Focus on Interpretable Models, Post-hoc Explanation, and Discrimination Testing}}%Attention AE/ME. The following layout issues have not been checked by the English Editing Department and must be carefully verified by the AE/Layout Department: All callout issues, bold usage of callouts, and references to callouts in the text. Correct callout usage in figures. Figure and Table layout issues. Footnote formatting and Glossaries have not been checked. En dash usage for negative values, en dash usage to indicate relationships, en dash usage to indicate bonds (especially in chemistry). The English Editing Department is not responsible for correct italic usage for genes, proteins and technical terminology. This responsibility belongs to the authors. The following are also not checked: spacing between numbers and units of measurement, ratios, en dashes for ranges, date and time formats, punctuation in equation lines, and less than/more than spacing (< >). Finally, capitalization and layout of titles/headings must be properly checked as well as ensuring 'Eq.' and 'Fig.' are properly spelled out, as these are layout issues.

% Authors, for the paper (add full first names)
\Author{\hl{Navdeep Gill} $^{1, \dagger}$, Patrick Hall $^{1,2,}$*$^{,\dagger}$, Kim Montgomery $^{1,\dagger}$ and Nicholas Schmidt $^{3,}$*$^{,\dagger}$} %%%Author order is not the same as on the system, please confirm which one is correct.
%Please carefully check the accuracy of names and affiliations.

% Authors, for metadata in PDF
\AuthorNames{Navdeep Gill, Patrick Hall, Kim Montgomery and Nicholas Schmidt}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad \hl{H2O.ai}, 2307 Leghorn St., Mountain View, CA, 94043, USA; navdeep.gill@h2o.ai (N.G.); kim.montgomery@h2o.ai (K.M.)\\%%%Please provide the department, university,zip code, city, country.
$^{2}$ \quad \hl{The George Washington University}, Department of Decision Sciences, 2201 G Street NW,
Washington, D.C., 20052, USA\\%%%Please provide the department, zip code, city, country.
$^{3}$ \quad \hl{BLDS, LLC}, 1608 Walnut St., Suite 1108, Philadelphia, PA, 19103, USA}%%%Please provide the department, university,zip code, city, country.

% Contact information of the corresponding author
\corres{Correspondence: phall@h2o.ai (P.H.); nschmidt@bldsllc.com (N.S.)}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines and explainable neural networks, a deep learning architecture well-suited for structured data, are assessed on simulated data and publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence and individual conditional expectation and with global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact and other types of discrimination using measures with long-standing legal precedents, adverse impact ratio, marginal effect, and standardized mean difference, along with straightforward group fairness measures. By combining interpretable models, post-hoc explanations, and discrimination testing with accessible software tools, this text aims to provide a template workflow for machine learning applications that require high accuracy and interpretability and that mitigate risks of discrimination.}

% Keywords
\keyword{deep learning; disparate impact; explanation; fairness; gradient boosting machine; interpretable; machine learning; neural network; Python}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{section}{0}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that~perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds ({\hl{See}%%% footnote is not allowed in our jounral, we have moved them in the main text, please confirm if you are OK with this. 
%is the text color necessary? please confirm if need remove the link of the text,
%% we have removed the link of text, and keep web link
: 
% THESE HYPERLINKS ARE NECESSARY - PH
\href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}
{Responsible Artificial Intelligence}, 
\href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}
{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, 
\href{https://www.pwc.com/gx/en/issues/data-and-analytics/artificial-intelligence/what-is-responsible-ai.html}
{PwC's Responsible AI}, 
\href{https://ai.google/responsibilities/responsible-ai-practices/}
{Responsible AI Practices}). Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is becoming clear. Machine learning (ML) models, a common application of AI, can present serious risks. ML~models can be inaccurate and unappealable %INCORRECT - PH -- Please confirm that the intended meaning has been retained.
black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop} ({\hl{See}% footnote is not allowed in our jounral, we have moved them in the main text, please confirm if you are OK with this. 
: 
\href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}
{\textit{When a Computer Program Keeps You in Jail}}). ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying,dwork2012fairness,gender_shades}, and ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml,model_stealing,membership_inference,shokri2019privacy}. This~manuscript makes no claim that these interdependent issues of ML opaqueness, discrimination, privacy harms, and security vulnerabilities have been resolved, even as singular entities, and much less as complex intersectional phenomena. However, Sections \ref{sec:m_and_m}--\ref{sec:disc} do propose some specific technical countermeasures, mostly in the form of interpretable models, post-hoc explanation, and disparate impact (DI) and discrimination testing, that responsible practitioners can use to address a subset of these vexing problems ({\hl{This text and associated software are not, and should not be construed as, legal advice or regulatory compliance requirements}}), ({\hl{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve Supervision and Regulation (SR) Letter 11-7, the European Union (EU) General Data Protection Regulation (GDPR) Article} 22, or other laws and regulations \cite{ff_interpretability}.})%%% footnote is not allowed in our jounral, we have moved them in the main text, please confirm if you are OK with this.

Section \ref{sec:m_and_m} describes methods and materials, including training datasets, interpretable and constrained models, post-hoc explanations, tests for DI and other social discrimination, and public and open source software resources associated with this text. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models, and~post-hoc explanation and discrimination testing results are also presented for interpretable models. Of course, an even wider array of tools and techniques are likely helpful to fully minimize discrimination, inaccuracy, privacy, and security risks associated with ML models. Section \ref{sec:disc} puts forward a more holistic responsible ML modeling workflow, and addresses the burgeoning Python ecosystem for responsible AI, along with appeal and override of automated decisions, and discrimination testing and remediation in practice. Section \ref{sec:con} closes this manuscript with a brief summary of the outlined methods, materials, results, and discussion.

@@

The simulated data (see Section~\ref{ssec:sim_data}) are based on the well-known Friedman datasets. Its~known feature importance and augmented discrimination characteristics are used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1979tree,friedman1991multivariate}. The~mortgage data (see Section~\ref{ssec:mort_data}) are sourced from the Home Mortgage Disclosure Act (HMDA) database, a fairly realistic data source for demonstrating the template workflow \cite{hmda} ({\hl{See:} 
\href{https://www.consumerfinance.gov/data-research/hmda/}
{Mortgage data (HMDA})}). %%% footnote is not allowed in our jounral, we have moved them in the main text, please confirm if you are OK with this. 
To provide a sense of fit differences, performance is compared on simulated data and collected mortgage data between the more interpretable constrained ML models and the less interpretable unconstrained ML models. Because the unconstrained ML models, gradient boosting machines (GBMs, e.g., \cite{friedman2001greedy,esl}) and artificial neural networks (ANNs, e.g., \cite{recht2011hogwild,hinton2012improving,sutskever2013importance,zeiler2012adadelta}), do not exhibit convincing accuracy benefits on the simulated or mortgage data and can also present the unmitigated risks discussed above, further explanation and discrimination analyses are applied only to the constrained, interpretable ML models \cite{please_stop,fair_washing,scaffolding}. Here, monotonic gradient boosting machines ({\hl{as implemented in} 
%\href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}
{\texttt{\hl{XGBoost}}} or 
%\href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}
{\texttt{\hl{h2o}}}.}) %% footnote is not allowed in our jounral, we have moved them in the main text, please confirm if you are OK with this. % we have removed the link of text, and keep web link
(MGBMs, see Section~\ref{ssec:mgbm}) and explainable neural networks (XNNs, see Section~\ref{ssec:xnn} \cite{,wf_xnn,yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analyses. MGBM and XNN interpretable model architectures are selected for the example workflow because they are straightforward variants of popular unconstrained ML models. If~practitioners are working with GBM and ANN models, it should be relatively uncomplicated to also evaluate the constrained versions of these models.

The same can be said of the selected explanation methods and discrimination tests. Due to their post-hoc nature, they can often be shoe-horned into existing ML workflows and pipelines. Presented explanation techniques include partial dependence (PD) and individual conditional expectation (ICE) (see Section~\ref{ssec:pd_ice}) and Shapley values (see Section~\ref{ssec:shap}) \cite{esl,ice_plots,shapley,tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discrimination testing methods discussed (see Section~\ref{ssec:di}) include adverse impact ratio (AIR), marginal effect (ME), and standardized mean difference (SMD)  \cite{feldman2015certifying,cohen1988statistical,cohen1992power}. ({\hl{See: Part 1607 - Uniform Guidelines on Employee Selection Procedures} (1978) \textsection 1607.4.}).
%%% footnote is not allowed in our jounral, we have moved them in the main text, please confirm if you are OK with this. 
 Accuracy and other confusion matrix measures are also reported by demographic segment \cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available on GitHub (see Section~\ref{ssec:soft}).

@@

Simulated data are created based on a signal-generating function, $f$, applied to input data, $\mathbf{X}$, first~proposed in \citet{friedman1979tree} and extended in \citet{friedman1991multivariate}:
\begin{equation}
\begin{aligned}\label{eq:sim}\small
f(\mathbf{X}) = 10\sin(\pi~X_{\text{Friedman},1}~X_{\text{Friedman},2}) + 20(X_{\text{Friedman},3} - 0.5)^2 + 10~X_{\text{Friedman},4} + 5~X_{\text{Friedman},5}
\end{aligned}
\end{equation}
where each $X_{\text{Friedman},j}$ is a random uniform feature in $[0,1]$. In Friedman's texts, a Gaussian noise term was added to create a continuous output feature for testing spline regression methodologies. In this manuscript, the signal-generating function and input features are modified in several ways.  Two binary features, a categorical feature with five discrete levels, and a bias term are introduced into $f$ to add a degree of complexity that may more closely mimic real-world settings. 
%When generating the data, random draws for each observation are constructed such that the features can be correlated with each other, some highly so. For example, X1 and X3 have a Pearson correlation coefficient equal to 0.58.
For binary classification analysis, the Gaussian noise term is replaced with noise drawn from a logistic distribution and coefficients are re-scaled to be $\frac{1}{5}$ of the size of those used by Friedman, and any $f(\mathbf{X})$ value above 0 is classified as a positive outcome, while $f(\mathbf{X})$ values less than or equal to zero are designated as negative outcomes.  Finally, $f$ is augmented with two hypothetical protected class-control features with known dependencies on the binary outcome to allow for discrimination testing. The simulated data are generated to have eight input features, twelve after numeric encoding of categorical features, and a binary outcome, two class-control features, and 100,000 instances. The simulated data are then split into a training and test set, with 80,000 and 20,000 instances, respectively.  Within the training set, a~five-fold cross-validation indicator is used for training all models. For an exact specification of the simulated data, see the software resources referenced in Section~\ref{ssec:soft}.

@@

The mortgage dataset analyzed here is a random sample of consumer-anonymized loans from the HDMA database. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e., the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  After data cleaning and preprocessing to encode categorical features and create missing markers, the mortgage data contain ten input features and the binary outcome, \textit{high-priced}. The data are split into a training set with 160,338 loans and a marker for 5-fold cross-validation and a test set containing 39,662 loans. While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (loan to value (LTV) ratio, debt to income (DTI) ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider. See the resources put forward in Section \ref{ssec:soft} and Appendix \ref{a_sec:mort} for more information regarding the HMDA mortgage data.

@@

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints, with respect to each input feature, $X_j$, and a target feature, $\mathbf{y}$, independently. An MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, and each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$. For an instance, $\mathbf{x}$, a trained MGBM model, $g^{\text{MGBM}}$, takes the form:

@@

 As in unconstrained GBM, $\Theta^{\text{mono}}_b$ is selected in a greedy, additive fashion by minimizing a regularized loss function that considers known target labels, the predictions of all subsequently trained trees in $g^{\text{MGBM}}$, and the $b$-th tree splits applied to $\mathbf{x}$, $T_b(\mathbf{x};\Theta^\text{mono}_b)$, in a numeric loss function (e.g., squared loss, Huber loss), and a regularization term that penalizes complexity in the current tree. See~Appendices \ref{a_ssec:not} and \ref{a_ssec:mgbm} for details pertaining to MGBM training.

Herein, two $g^\text{MGBM}$ models are trained. One on the simulated data and one on the mortgage data. In both cases, positive and negative monotonic constraints for each $X_j$ are selected using domain knowledge, random grid search is used to determine other hyperparameters, and five-fold cross-validation and test partitions are used for model assessment. For exact parameterization of the two $g^\text{MGBM}$ models, see the software resources referenced in Section~\ref{ssec:soft}.

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bear a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}Ms), which consider main effects and a small number of two-way interactions and may also incorporate boosting into their training \cite{esl,ga2m}. XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. A trained XNN function, $g^{\text{XNN}}$, applied to some instance , $\mathbf{x}$, is defined as:
\begin{equation}
\begin{aligned}
\label{eq:xnn}
g^{\text{XNN}}(\mathbf{x}) = \mu_0 + \sum_{k=0}^{K-1}\gamma_k n_k(\sum^{J-1}_{j=0}\beta_{k, j}x_j)
\end{aligned}
\end{equation}
where $\mu_0$ is a global bias for $K$ individually specified ANN subnetworks, $n_k$, with weights $\gamma_k$. The~inputs to each $n_k$ are themselves a linear combination of the $J$ modeling inputs and their associated $\beta_{k,j}$ coefficients in the deepest, i.e., {projection}, layer of $g^\text{XNN}$.

Two $g^\text{XNN}$ models are trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in five training folds and in a test data partition. $L_1$ regularization is applied to network weights to induce a sparse and interpretable model, where~each $n_k$ and corresponding $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Section~\ref{ssec:soft} and Appendices  \ref{a_ssec:not} and \ref{a_ssec:xnn}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One-Dimensional Partial Dependence and Individual Conditional Expectation}\label{ssec:pd_ice}

PD plots are a widely-used method for describing and plotting the estimated average prediction of a complex model, $g$, across some partition of data, $\mathbf{X}$, for some interesting input feature, $X_j \in \mathbf{X}$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ with regard to values of an input feature in a single instance, $x_j$. PD and ICE can be overlaid in the same plot to create a holistic global and local portrait of the predictions for some $g$ and $X_j$ \cite{ice_plots}. When $\text{PD}(X_j, g)$ and $\text{ICE}(x_j,  g)$ curves diverge, such plots can also be indicative of modeled interactions in $g$ or expose flaws in PD estimation, e.g., inaccuracy in the presence of strong interactions and correlations \cite{ice_plots,ale_plot}.  For details regarding the calculation of one-dimensional PD and ICE, see the software resources in Section~\ref{ssec:soft} and Appendices~\ref{a_ssec:not} and \ref{a_ssec:pd_ice}.
